---
title: "PLSC 30600 Lab 6"
pdf-engine: xelatex
format:
  html:
    toc: false
    number-sections: false
    colorlinks: true
mainfont: XCharter
monofont: Fira Sans
author: "Zikai Li"
date: last-modified
cache: false
linestretch: 1.5
header-includes:
  - "\\usepackage{amsmath}"
  - "\\usepackage{tikz}"
  - "\\usepackage{amssymb}"
---
```{r packages, message=F, warning=F}
options(digits=3)
library(tidyverse)
library(table1)
library(haven)
library(broom)
library(gtsummary)
library(estimatr)
library(printr, quietly=TRUE)
library(margins)
library(xtable)
library(kableExtra)
library(sysfonts)
```

## Review of assumptions for IV estimation 

## Survey experiments: What are you trying to manipulate 

In the context of survey experiments, we are often interested in estimating the causal effect of a treatment on an outcome. The treatment is often contained in a piece of text. The outcome is often a measure of attitudes, beliefs, or behaviors.

Are we interested in the effect of the *text* itself, or the effect of some underlying *construct* that the text is meant to manipulate?

## The role of the manipulation check
The (treatment-relevant) manipulation check is a measure of the extent to which the treatment has the intended effect on the construct of interest (Kane and Barabas 2019).

It is usually a multiple choice question that asks respondents to recall the treatment. 

## Kim and Mutz (2017)

Mutz and Kim (2017) study the effect of "in-group favoritism" on trade preferences. They manipulate the expected gains from a trade deal for the US and the other country. 

- Manipulation of theoretical interest: perception of relative gain for US from trade deal
- Downstream outcome of interest: support for trade deal

*There is much concern these days about intentional trade and job security. We are going to describe a situation the United States could face in the future. Some parts of the description may strike you as important; other parts may seem unimportant. Please read the details very carefully. After describing the situation, we will ask your opinion about a policy option. Here is the situation: The United States is considering a trade policy that would have the following effects:*

*For each 1,000 / 10 people in the U.S. who gain a job and can now provide for their family, 10 / 1000 people in a country that we trade with will gain new jobs and now be able to provide for their family / lose jobs and will no longer be able to provide for their family.*

The manipulation check asks respondents to recall whether the US gains relative to the other country. The authors argue that Americans have a tendency to favor jobs for Americans, sometimes at the expense of jobs for other countries ("intergroup competition" mentality). 

### Treatment coding

BKRTW's coding:

| Treatment arm | Varying info |
| --- | --- |
| Baseline | US gains 10 jobs, other country gains 1,000 |
| Baseline | US gains 10 jobs, other country loses 1,000 |
| Treatment | US gains 1,000 jobs, other country gains 10 |


Our version:

| Treatment arm | Varying info |
| --- | --- |
| Baseline | US gains 10 jobs, other country gains 1,000 |
| Treatment | US gains 10 jobs, other country loses 1,000 |
| Treatment | US gains 1,000 jobs, other country gains 10 |

### The manipulation check
Think back to the trade policy that was described to you earlier in the survey. Will our trading partner benefit more than the US, will the US benefit more than the trading partner, or will they be impacted equally? 

Possible responses include: 

- The trading partner will benefit more than the US 
- The US will benefit more than trading 
- Both countries will benefit equally


```{r}
study1 <- read_csv("analysis_study1_ajps.csv")

mk_data_fac <- read_rds("mk_data.rds")
study1$mk_dv <- scale(as.numeric(study1$mk_dv), 
                      center = TRUE, scale = TRUE)

mk1 <- study1 %>% 
 filter(.,
        mk_context_fac == "Original") %>% 
 lm_robust(mk_dv ~ mk_main_treat, data = .) %>% 
 tidy(.) %>% filter(term!= "(Intercept)") %>% 
 mutate(., 
        term = "Trade Preference:\nITT Effect \n on Support \nfor Trade",
        Model = "Replication")

mk2 <- mk1 %>% rbind(study1 %>% 
 filter(.,
        mk_context_fac == "Original") %>% 
 iv_robust(mk_dv ~ I(mk_correct_recall*mk_main_treat)|mk_main_treat, data = ., se = "HC0") %>% 
 tidy(.) %>% filter(term!= "(Intercept)") %>% 
 mutate(., 
        term = "Trade Preference: \nIV Effect of Info Reception \n on Support \nfor Trade",
        Model = "IV"))
```

## Comparisons of results
```{r}

mk3 <- mk2 %>% rbind(study1 %>%
 filter(.,
        mk_context_fac == "Original", (mk_correct_recall ==0)) %>%
 lm_robust(mk_dv ~ mk_main_treat, data = .) %>%
 tidy(.) %>% filter(term!= "(Intercept)") %>%
 mutate(.,
        term = "Trade Preference:\n US Expected Gain \nIV Effect \n on Support \nfor Trade",
        Model = "Replication"))
```

Placebo test 
```{r placebo_test_mk, message=F}
# ----------------------------------------------------------------------
# Read data
# ----------------------------------------------------------------------

mk_data_fac <- readRDS("mk_data.rds")

# ----------------------------------------------------------------------
# Define and re-code main variables
# ----------------------------------------------------------------------

mk_data_fac_transf <- 
  mk_data_fac %>% 
 mutate(.,
        # Create factor variable of treatment conditions
        mk_lose_win_fac = case_when(startsWith(winlose, "For each 1,000 people in the U.S.") ~ "US Gains More",
                                    startsWith(winlose, "For each 10 people in the U.S. who gain a job and can now provide for their family, 1000 people in a country that we trade with will gain") ~ "US Gains Less",
                                    startsWith(winlose, "For each 10 people in the U.S. who gain a job and can now provide for their family, 1000 people in a country that we trade with will lose") ~ "Other Country loses"
        ),
        # create factor of context conditions
        mk_context_fac = case_when(
         is.na(mutz_kim_addition1) ~ "Original",
         mutz_kim_addition1 == "If approved, this policy will be implemented within the next two years. As part of the implementation process, a commission of government officials and bureaucrats will outline the financial implications of the policy and provide guidance to businesses on how the new agreement affects them. Lastly, a team comprised of bureaucrats from both countries will oversee the policy implementation process which is expected to last two years." ~ "Extra Innocuous",
         mutz_kim_addition1 == "If approved, this policy will be implemented within the next two years. Analysis of the agreement has determined that it will dramatically increase trade between the countries. This has the potential to create new business opportunities in both countries, but may also make it harder for some companies to compete. Lastly, a team comprised of bureaucrats from both countries will oversee the policy implementation process which is expected to last two years." ~ "Extra Non-Innocuous"
        )) 


### Edit Main Outcomes and Treatments for MK study in our replication

mk_data_fac_transf1 <- 
  mk_data_fac_transf %>% 
 mutate(
  # Create 1-4 scale of outcome
  mk_dv = case_when(
   mk_support_extent == 2 ~ 3,
   mk_support_extent == 1 ~ 4,
   mk_oppose_extent == 1 ~ 1,
   mk_oppose_extent == 2 ~ 2
  ),
  # Create main treatment condition, with relative gain as one category and
  # the other two as a category
  mk_main_treat = ifelse(mk_lose_win_fac == "US Gains Less",
                         0,
                         1),
  # Brutger et al.'s original treatment assignment indicator
    us_gains_more = ifelse(mk_lose_win_fac == "US Gains More",
                         1,
                         0),
  other_loses = ifelse(mk_lose_win_fac == "Other Country Loses",
                         1,
                         0)) 



# Treatment recall (authors' coding)
# Values for mk_post_treatment_2: 
# 1) Other country gains more
# 2) U.S. gains more
# 3) Both countries gain equally
# I code correct recall such that: 0 = wrong answer, 1 = correct answer
mk_data_fac_transf2 <- 
  mk_data_fac_transf1 %>% 
 mutate(.,
        mk_correct_recall = case_when(
         mk_lose_win_fac == "US Gains More" & mk_post_treatment_2 == 2 ~ 1,
         mk_lose_win_fac == "US Gains More" & mk_post_treatment_2 != 2 ~ 0,
         mk_lose_win_fac == "Other Country loses" & mk_post_treatment_2 == 2 ~ 1,
         mk_lose_win_fac == "Other Country loses" & mk_post_treatment_2 != 2 ~ 0,
         mk_lose_win_fac == "US Gains Less" & mk_post_treatment_2 == 1 ~ 1,
         mk_lose_win_fac == "US Gains Less" & mk_post_treatment_2 != 1 ~ 0
        ))

mk_data_fac_transf2$mk_dv <- scale(as.numeric(mk_data_fac_transf2$mk_dv), 
                      center = TRUE, scale = TRUE)

```

## Placebo test
Restrict reduced-form analysis to those in placebo/treatment that failed the manipulation check
```{r mk estimation}
mk1 <- mk_data_fac_transf2 %>%
                       filter(.,
                              mk_context_fac == "Original", (mk_correct_recall ==0)) %>%
                       lm_robust(mk_dv ~ us_gains_more, data = .) %>%
                       tidy(.) %>% filter(term!= "(Intercept)") %>%
                       mutate(.,
                              term = "Original Assignment \nIndicator and Support \nfor Trade Deal",
                              Model = "Placebo test")
# Restrict reduced-form analysis to those in placebo/treatment that failed the manipulation check
mk2 <- mk1 %>% rbind(  mk_data_fac_transf2 %>% 
  filter(.,
         mk_context_fac == "Original") %>% 
  lm_robust(mk_dv ~ us_gains_more, data = .) %>% 
  tidy(.) %>% filter(term!= "(Intercept)") %>% 
  mutate(., 
         term = "Original Assignment \nIndicator and Support \nfor Trade Deal",
         Model = "ITT"))

# altered treatment assignment indicator
mk3 <- mk2 %>% 
  rbind(mk_data_fac_transf2 %>%
                       filter(.,
                              mk_context_fac == "Original", (mk_correct_recall ==0)) %>%
                       lm_robust(mk_dv ~ mk_main_treat, data = .) %>%
                       tidy(.) %>% filter(term!= "(Intercept)") %>%
                       mutate(.,
                              term = "Altered Assignment \nIndicator and Support \nfor Trade Deal",
                              Model = "Placebo test")) %>% rbind(mk_data_fac_transf2 %>% 
  filter(.,
         mk_context_fac == "Original") %>% 
  lm_robust(mk_dv ~ mk_main_treat, data = .) %>% 
  tidy(.) %>% filter(term!= "(Intercept)") %>% 
  mutate(., 
         term = "Altered Assignment \nIndicator and Support \nfor Trade Deal",
         Model = "ITT")) 


ggplot(mk3, aes(x = term, y = estimate, color = Model)) +
  geom_hline(yintercept = 0, color = "gray50", linetype = 2) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high,
                      color = Model, shape = Model),
                  position = position_dodge(width = 0.4),
                  fill = "white") +
  scale_color_manual(values = c("#3B9AB2", "#E1AF00"))+
  guides(col = guide_legend(reverse = F),
         shape = guide_legend(reverse = F))+
  #ylim(-0.1,1.2)+
  coord_flip()+
  labs(x = "",
       y = "Estimates with 95% Confidence Intervals") +
  scale_x_discrete(limits = rev(unique(mk3$term))) +
  theme(text = element_text(size = 12, family = "XCharter"),
        legend.key=element_blank(),
        panel.grid.major = element_blank(), 
        axis.text.x = element_text(size = 12, color = "black"),
        axis.text.y = element_text(colour="black"),
        plot.caption = element_text(size = 12,hjust = -.02),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) -> mk_ate_vs_placebo_test_plt

print(mk_ate_vs_placebo_test_plt)

mk_data_fac_transf3 <- mk_data_fac_transf2 %>%
  mutate(mk_lose_win_fac = factor(mk_lose_win_fac, ordered = F)) %>% 
  mutate(mk_lose_win_fac = relevel(mk_lose_win_fac, ref= "US Gains More"))

mk4 <-  mk_data_fac_transf3 %>%
   filter(.,
          mk_context_fac == "Original") %>%
   lm_robust(mk_dv ~ mk_lose_win_fac, data = .) %>%
   tidy(.) %>% filter(term!= "(Intercept)") %>%
   mutate(.,
          Term = case_when(
          term == "mk_lose_win_facOther Country loses" ~
            "US: +10, Other: -1000",
          term == "mk_lose_win_facUS Gains Less" ~
            "US: +10; Other: +1000",
            TRUE ~ term),
          Model = "ITT") %>% 
  rbind(mk_data_fac_transf3 %>% 
  filter(.,
        mk_context_fac == "Original", mk_correct_recall ==0 ) %>%
  lm_robust(mk_dv ~ mk_lose_win_fac, data = .) %>%
  tidy(.) %>% filter(term!= "(Intercept)") %>%
  mutate(.,
        Term = case_when(
          term == "mk_lose_win_facOther Country loses" ~
            "US: +10, Other: -1000",
          term == "mk_lose_win_facUS Gains Less" ~
            "US: +10; Other: +1000",
          TRUE ~ term),
        Model = "Restricted Sample: \nIncorrect Recall")) 
  
  
mk5_correct_recall  <-   (mk_data_fac_transf3 %>% 
  filter(.,
        mk_context_fac == "Original", mk_correct_recall ==1) %>%
  lm_robust(mk_dv ~ mk_lose_win_fac, data = .) %>%
  tidy(.) %>% filter(term!= "(Intercept)") %>%
  mutate(.,
        Term = case_when(
          term == "mk_lose_win_facOther Country loses" ~
            "US: +10, Other: -1000",
          term == "mk_lose_win_facUS Gains Less" ~
            "US: +10; Other: +1000",
          TRUE ~ term),
        Model = "Restricted Sample: \nCorrect Recall"))



ggplot(mk4, aes(x = Model, y = estimate, color = Term)) +
  geom_hline(yintercept = 0, color = "gray50", linetype = 2) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high,
                      color = Term, shape = Term),
                  position = position_dodge(width = 0.4),
                  fill = "white") +
  scale_color_manual(values = c("#3B9AB2", "#E1AF00"))+
  guides(col = guide_legend(reverse = F),
         shape = guide_legend(reverse = F))+
  #ylim(-0.1,1.2)+
  coord_flip()+
  labs(x = "",
       y = "Estimates with 95% Confidence Intervals"#,
       #title = "Factorial assignment indicator \nBaseline: US: +1000; Other: +10"
       ) +
  scale_x_discrete(limits = rev(unique(mk4$Model))) +
  theme(text = element_text(size = 12, family = "XCharter"),
        legend.key=element_blank(),
        panel.grid.major = element_blank(), 
        axis.text.x = element_text(size = 12, color = "black"),
        axis.text.y = element_text(colour="black"),
        plot.caption = element_text(size = 12,hjust = -.02),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(), 
        axis.line = element_line(colour = "black")) -> mk_ate_vs_placebo_test_plt2


print(mk_ate_vs_placebo_test_plt2)

```



## References
Brutger, Ryan, Joshua D. Kertzer, Jonathan Renshon, Dustin Tingley, and Chagai M. Weiss. "Abstraction and detail in experimental design." *American Journal of Political Science* 67, no. 4 (2023): 979-995.

Kane, John V., and Jason Barabas. "No harm in checking: Using factual manipulation checks to assess attentiveness in experiments." *American Journal of Political Science* 63, no. 1 (2019): 234-249.

Li, Zikai, and Robert Gulotty. "Information Exposure and Belief Manipulation in Survey Experiments." *Woking Paper* (2023).

Mutz, Diana C., and Eunji Kim. "The impact of in-group favoritism on trade preferences." *International Organization* 71, no. 4 (2017): 827-850.
