---
title: 'Replication: Grumbach and Hill (2022) JoP'
author: ''
date: "8/25/2023"
output:
  pdf_document: default
  html_document: default
  word_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Required packages
library(tidyverse)
library(haven)
library(plyr)

# Quick regression
library(estimatr)

# Visualize distribution
library(panelView)

# CS DID 
library(did)

# Extras for plotting
library(gridExtra)
library(ggpubr)

# Print options
options(digits=3)

```

# Description of paper

Grumbach and Hill (2022) looks at how same day registration (SDR) laws affect turnout for different kinds of voters. It analyzes the effect of these laws on voter turnout in presidential and nonpresidential elections. The key outcome variable is voter turnout. The authors conduct subgroup analysis for different age groups.  

The full citation is

> Grumbach, Jacob M., and Charlotte Hill. "Rock the registration: Same day registration increases turnout of young voters." The Journal of Politics 84.1 (2022): 405-417.

|                       |                      |
|-----------------------|----------------------|
| Unit                  | Individual (primary) |
| Time                  | Year                 |
| Type of design        | Staggered rollout    |
| Replicated Figure(s)  | Figures A2           |
  
# Data description and pre-processing
  
The relevant variables are:
  
- `year` - Year
- `state` - State
- `voted` - Whether an individual voted in a given year
- `sdr` - Whether the state has same-day registration law; treatment indicator
- `presidential_year` - Whether there is a presidential election
- `age_group` - Age group of the individual



Data-processing using replication code from original paper
```{r, message=F, results='hide', warning=F, echo=T}
# State FIPS codes
fips_codes <- read_csv("fips_codes_website.csv")

data <- readRDS("policy_data_updated.RDS")

# Load New Early Voting Data
ev <- read.csv("Early Voting Coding - Sheet1.csv")

data$early_voting_narrow
data$early_voting_broad

data <- plyr::join(data, ev[,c("state","year_eip","year_amv","year_abs")])

for(i in levels(factor(data$state))){

  data$early_voting_person[data$state==i &
                             (data$year>=data$year_eip) &
                             data$year<=2014] <- 1

  data$early_voting_narrow[data$state==i &
                             (data$year>=data$year_eip|data$year>=data$year_amv) &
                             data$year<=2014] <- 1
  data$early_voting_broad[
    data$state==i &(data$year>=data$year_eip|data$year>=data$year_amv|
                      data$year>=data$year_abs) &
      data$year<=2014] <- 1

}

data$early_voting_narrow[is.na(data$early_voting_narrow) & data$year<=2014] <- 0
data$early_voting_broad[is.na(data$early_voting_broad) & data$year<=2014] <- 0
data$early_voting_person[is.na(data$early_voting_person) & data$year<=2014] <- 0

presidential_years <- c(1976,1980,1984,1988,1992,1996,2000,2004,2008,2012,2016)

cps <- read_csv("cps_00021.csv", col_types = cols(
  .default = col_double()
))
names(cps) <- tolower((names(cps)))

# Remove "refused to answer"
cps$voted_alt <- cps$voted - 1
cps$voted_alt[cps$voted_alt>1] <- NA

# Recode dependent variable
cps$voted[cps$voted==99] <- NA
cps$voted <- cps$voted - 1
cps$voted[cps$voted>1] <- 0

# Recode covariates
cps$faminc[cps$faminc>843] <- NA
cps$faminc <- as.numeric(factor(cps$faminc))

cps$sex[cps$sex>2] <- NA

cps$educ[cps$educ==999|cps$educ==1] <- NA

cps$educ <- as.numeric(factor(cps$educ))

#merge state FIPS codes
cps <- join(cps, fips_codes[,c("st","statefip")], match="first")

#Merge state policy data
cps <- join(cps, data, by=c("st","year")); gc()

cps$age_group[cps$age<=24] <- "18-24"
cps$age_group[cps$age>24 & cps$age<35] <- "25-34"
cps$age_group[cps$age>34 & cps$age<45] <- "35-44"
cps$age_group[cps$age>44 & cps$age<55] <- "45-54"
cps$age_group[cps$age>54 & cps$age<65] <- "55-64"
cps$age_group[cps$age>64] <- "65+"

cps$presidential_year[cps$year %in% presidential_years] <- 1
cps$presidential_year[(cps$year %in% presidential_years)==F] <- 0

cps$sex[cps$sex==0] <- NA
cps$sex <- cps$sex - 1

cps$race_5[cps$race==100] <- "White"
cps$race_5[cps$race==200] <- "Black"
cps$race_5[cps$race==300 | cps$race>=700] <- "Other"
cps$race_5[(cps$race>=600 & cps$race<700) | cps$race==809] <- "Asian"
cps <- within(cps, race_5 <- relevel(factor(race_5), ref = "White"))
cps <- cps[cps$year>=1978,]

cps$asian[cps$race_5=="Asian"] <- 1
cps$asian[cps$race_5!="Asian"] <- 0


gc()

# age_group dummy variables
cps <- cbind(cps, data.frame(model.matrix(~ age_group - 1, data=cps)))

# race dummy variables
cps <- cbind(cps, data.frame(model.matrix(~ as.character(race) - 1, data=cps)))
names(cps) <- gsub("as.character.", "", names(cps))

# registration
cps$reg_pollingplace[cps$voreghow<99 & cps$voreghow!=7] <- 0
cps$reg_pollingplace[cps$voreghow<99 & cps$voreghow==7] <- 1

cps$vote_reg[cps$voreg<=98 & cps$voreg!=2] <- 0
cps$vote_reg[cps$voreg<=98 & cps$voreg==2] <- 1


gc()

# Drop those with duration of residence less than 1 year
cps$movedless1yr[cps$voteres<14] <- 1
cps$movedless1yr[cps$voteres>=14 & cps$voteres<900] <- 0

# Remove DC
cps <- cps[cps$st!="DC",]; gc()

# Remove North Dakota since it gets dropped in all analyses b/c NA SDR
# This step is not in the original paper's code but it doesn't change results
cps <- cps[cps$st!="ND",]; gc()

paneldata <- data[data$year>=1978,
                  c("st","year","sdr","early_voting_narrow","early_voting_broad",
                    "early_voting_person", "pop_annual")]
paneldata$st <- as.character(paneldata$st)
```


# Treatment visualization

```{r, message=F, warning=F, results='hide', fig.height=8, fig.width=7}
# pdf("PanelView - SDR (new).pdf", h=7, w=12)
panelview(D="sdr", 
          data=data,
          #D="sdr",
          #by.timing = T,
          pre.post = TRUE,
          #data=paneldata,
          index=c("st","year"),
          xlab="Year",
          ylab="State",
          main="SDR",
          axis.adjust=T, 
          legend.labs = c("Control Observations", "Treated Observations"),
          color=c("white","black"),
          theme.bw = T)
```

# Exact replication

```{r, warning=F, results="hide", message=F}

### From original replication code
agg_cps_age <- aggregate(voted ~ state + year + sdr + presidential_year + 
                           age_group, cps, mean, na.rm=T)

agg_cps_age <- agg_cps_age[order(agg_cps_age$state, agg_cps_age$year),]

firstyears <- do.call("rbind", as.list(by(agg_cps_age[agg_cps_age$sdr==1,],
                                          agg_cps_age$state[agg_cps_age$sdr==1],
                                          head, n=1)))

firstyears$yearSDR <- firstyears$year

agg_cps_age <- join(agg_cps_age, firstyears[,c("state", "yearSDR")])


# Make a variable for how many years away a unit is from its treatment time 
# (0 is the time of treatment initiation, -1 is the time *just* before treatment)
agg_cps_age <- agg_cps_age %>% mutate(yearFromSDR = year - yearSDR)


# ever-treated variable
agg_cps_age$ever_treated[is.na(agg_cps_age$yearFromSDR)] <- 0
agg_cps_age$ever_treated[!is.na(agg_cps_age$yearFromSDR)] <- 1

# all never-treateds' yearFromSDR get assigned '0' 
agg_cps_age$yearFromSDR[is.na(agg_cps_age$yearFromSDR)] <- 0

cps_exact <- join(cps, agg_cps_age[agg_cps_age$age_group=="18-24",
                                   c("state", "year", "yearFromSDR", 
                                     "ever_treated", "yearSDR")])
gc()

cps_exact$yearFromSDR <- as.numeric(cps_exact$yearFromSDR)

# Make the dummy variables using the factor syntax - make -2 the reference period
cps_exact$yearFromSDRFactor <- relevel(as.factor(cps_exact$yearFromSDR), ref="-2")


# Make the dummy variables manually
cps_exact$`treat-16` <- as.numeric(cps_exact$yearFromSDRFactor == '-16')

cps_exact$`treat-14` <- as.numeric(cps_exact$yearFromSDRFactor == '-14')

cps_exact$`treat-12` <- as.numeric(cps_exact$yearFromSDRFactor == '-12')

cps_exact$`treat-10` <- as.numeric(cps_exact$yearFromSDRFactor == '-10')

cps_exact$`treat-8` <- as.numeric(cps_exact$yearFromSDRFactor == '-8')

cps_exact$`treat-6` <- as.numeric(cps_exact$yearFromSDRFactor == '-6')

cps_exact$`treat-4` <- as.numeric(cps_exact$yearFromSDRFactor == '-4')

cps_exact$`treat-2` <- as.numeric(cps_exact$yearFromSDRFactor == '-2')

cps_exact$`treat0` <- as.numeric(cps_exact$yearFromSDRFactor == '0')

cps_exact$`treat2` <- as.numeric(cps_exact$yearFromSDRFactor == '2')

cps_exact$`treat4` <- as.numeric(cps_exact$yearFromSDRFactor == '4')

cps_exact$`treat6` <- as.numeric(cps_exact$yearFromSDRFactor == '6')

cps_exact$`treat8` <- as.numeric(cps_exact$yearFromSDRFactor == '8')

cps_exact$`treat10` <- as.numeric(cps_exact$yearFromSDRFactor == '10')

cps_exact$`treat12` <- as.numeric(cps_exact$yearFromSDRFactor == '12')

cps_exact$`treat14` <- as.numeric(cps_exact$yearFromSDRFactor == '14')

cps_exact$`treat16` <- as.numeric(cps_exact$yearFromSDRFactor == '16')

cps_exact$include <- as.numeric(cps_exact$yearFromSDR <=16 & 
                                  cps_exact$yearFromSDR >=(-16))

# Code those with NAs in yearSDR as not having SDR
cps_exact$yearSDR_did <- ifelse(is.na(cps_exact$yearSDR), 0, cps_exact$yearSDR)
gc()
```

"Bivariate" results (This outputs an exact replication of Figure A2 in original paper)

```{r, fig.width=8, fig.height=5}
# By age-group
age_group_levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")

# Store results for each age group
age_group_results <- list()

# For each age group
for (group_level in age_group_levels){
  #Fit a regression to get the per-period tests

  dyn_reg_1 <- lm_robust(voted ~ `treat-16` +  `treat-14` +  `treat-12` +  
                           `treat-10`  + `treat-8`  + `treat-6` + `treat-4` + 
                           `treat0`  + `treat2`  + `treat4`  + `treat6` +  
                           `treat8` + `treat10` +`treat12`  + `treat14`  + 
                           `treat16` , fixed_effects = ~ as.factor(year) + 
                           as.factor(state), data=cps_exact[
                             cps_exact$age_group==group_level &
                               (cps_exact$yearFromSDR>=(-16) &
                                  cps_exact$yearFromSDR<=16) & 
                               !is.na(cps_exact$voted),])
  # Original paper clustered wrong, so don't do CRSEs
  
  # Make into a plotting set
  dyn_plot_1 <- tidy(dyn_reg_1) %>% 
    filter(!is.na(estimate)) %>% 
    mutate(term = str_remove_all(term, "`")) %>% 
    mutate(period = as.numeric(str_remove(term, fixed("treat")))) %>% 
    select(period, estimate, conf.low, conf.high)
  
  # Tack on the baseline
  dyn_plot_1 <- rbind(dyn_plot_1, data.frame(period = -2, estimate = 0, 
                                             conf.low = NA, conf.high = NA))
  
  # Label the age_group
  dyn_plot_1$age_group <- group_level
  
  # Append age_group_results
  age_group_results[[group_level]] <- dyn_plot_1
  
}

dyn_plot_combined <- bind_rows(age_group_results)

# Coefficient plot
exact_1 <- dyn_plot_combined %>% 
  ggplot(aes(x=as.numeric(period), y = estimate, ymin=conf.low, ymax=conf.high)) + 
  geom_ribbon(fill="grey80") +
  geom_line() +
  facet_wrap(~age_group) +
  geom_vline(xintercept = -0, lty=2) + geom_hline(yintercept = 0, lty=2) +
  xlab("Years Until SDR Implementation") + ylab("Treatment Effect")  + 
  scale_x_continuous(breaks=c(-16,-12,-8,-4,0,4,8,12,16)) +
  scale_y_continuous(breaks=c(-0.1,-0.05,0,0.05,0.1), limits=c(-.1,.1)) +
  theme_classic()

print(exact_1)
ggsave("figs/pretrend_by_age_exact_1.png",exact_1, width = 7, height = 4.3)
ggsave("figs/pretrend_by_age_exact_1.pdf",exact_1, width = 7, height = 4.3)
```

# Corrected replications

## Replication 1 (The replication figure we use in our paper)

Changes: 
- Clustered the SEs (the original paper's code contained a typo for clustered standard errors, which we correct here)

```{r, fig.width=8, fig.height=5}
# By age-group
age_group_levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")

# Store results for each age group
age_group_results <- list()

# For each age group
for (group_level in age_group_levels){
  #Fit a regression to get the per-period tests

  dyn_reg_1 <- lm_robust(voted ~ `treat-16` + `treat-14` + `treat-12` + 
                           `treat-10`  + `treat-8`  + `treat-6` + `treat-4` + 
                           `treat0`  + `treat2`  + `treat4`  + `treat6` +  
                           `treat8` + `treat10` +`treat12` + `treat14` + 
                           `treat16`, 
                         fixed_effects = 
                           ~ as.factor(year) + as.factor(state), 
                         data=cps_exact[
                             cps_exact$age_group==group_level &
                               (cps_exact$yearFromSDR>=(-16) &
                                  cps_exact$yearFromSDR<=16)&
                               !is.na(cps_exact$voted),], 
                         clusters=state, se_type="stata")
  
  #print(dyn_reg_1)
  
  # Make into a plotting set
  dyn_plot_1 <- tidy(dyn_reg_1) %>% filter(!is.na(estimate)) %>% 
    mutate(term = str_remove_all(term, "`")) %>% 
    mutate(period = as.numeric(str_remove(term, fixed("treat")))) %>% 
    select(period, estimate, conf.low, conf.high)
  
  # Tack on the baseline
  dyn_plot_1 <- rbind(dyn_plot_1, data.frame(period = -2, estimate = 0, 
                                             conf.low = NA, conf.high = NA))
  
  # Label the age_group
  dyn_plot_1$age_group <- group_level
  
  age_group_results[[group_level]] <- dyn_plot_1
  
}

dyn_plot_combined <- bind_rows(age_group_results)

# Coefficient plot
replication_1 <- dyn_plot_combined %>% 
  ggplot(aes(x=as.numeric(period), y = estimate, ymin=conf.low, ymax=conf.high)) + 
  geom_ribbon(fill="grey80") +
  geom_line() +
  facet_wrap(~age_group) +
  geom_vline(xintercept = -0, lty=2) + geom_hline(yintercept = 0, lty=2) +
  xlab("Years Until SDR Implementation") + ylab("Treatment Effect")  +
  scale_x_continuous(breaks=c(-16,-12,-8,-4,0,4,8,12,16)) +
  scale_y_continuous(breaks=c(-.2,-.15, -0.1,-0.05,0,0.05,0.1,0.15, 0.2), 
                     limits=c(-.2,.2)) +
  theme_classic()

print(replication_1)

ggsave("figs/pretrend_by_age_rep_1.png",replication_1, width = 7, height = 4.3)

ggsave("figs/pretrend_by_age_rep_1.pdf",replication_1, width = 7, height = 4.3)
```

## Replication 2 (First extension)

Changes
- Clustered the SEs
- Dropped the always-treated groups
- Correctly re-coded the never-treated groups

```{r, warning=F, results="hide", message=F}

### From original replication code
agg_cps_age <- aggregate(voted ~ state + year + sdr + presidential_year + age_group, 
                         cps, mean, na.rm=T)

agg_cps_age <- agg_cps_age[order(agg_cps_age$state, agg_cps_age$year),]

firstyears <- do.call("rbind", as.list(by(agg_cps_age[agg_cps_age$sdr==1,],
                                          agg_cps_age$state[agg_cps_age$sdr==1],
                                          head, n=1)))
firstyears$yearSDR <- firstyears$year

agg_cps_age <- join(agg_cps_age, firstyears[,c("state", "yearSDR")])

# Make a variable for how many years away a unit is from its treatment time 
# (0 is the time of treatment initiation, -1 is the time *just* before treatment)
agg_cps_age <- agg_cps_age %>% mutate(yearFromSDR = year - yearSDR)

# ever-treated variable
agg_cps_age$ever_treated[is.na(agg_cps_age$yearFromSDR)] <- 0
agg_cps_age$ever_treated[!is.na(agg_cps_age$yearFromSDR)] <- 1

# all never-treateds' yearFromSDR get assigned '0' 
agg_cps_age$yearFromSDR[is.na(agg_cps_age$yearFromSDR)] <- Inf

# Which ones are always treated
agg_cps_age$always_treated <- 0
agg_cps_age$always_treated[agg_cps_age$yearSDR == 1978] <- 1

# Ohio reverses
agg_cps_age$reversed <- 0
agg_cps_age$reversed[agg_cps_age$state == "Ohio"] <- 1

cps_rep2 <- join(cps, agg_cps_age[agg_cps_age$age_group=="18-24",
                                  c("state", "year", "yearFromSDR", 
                                    "ever_treated", "always_treated", 
                                    "reversed")])
gc()

cps_rep2$yearFromSDR <- as.numeric(cps_rep2$yearFromSDR)

# Make the dummy variables using the factor syntax - make -2 the reference period
cps_rep2$yearFromSDRFactor <- relevel(as.factor(cps_rep2$yearFromSDR), ref="-2")

cps_rep2$`treat-16` <- as.numeric(cps_rep2$yearFromSDRFactor == '-16')

cps_rep2$`treat-14` <- as.numeric(cps_rep2$yearFromSDRFactor == '-14')

cps_rep2$`treat-12` <- as.numeric(cps_rep2$yearFromSDRFactor == '-12')

cps_rep2$`treat-10` <- as.numeric(cps_rep2$yearFromSDRFactor == '-10')

cps_rep2$`treat-8` <- as.numeric(cps_rep2$yearFromSDRFactor == '-8')

cps_rep2$`treat-6` <- as.numeric(cps_rep2$yearFromSDRFactor == '-6')

cps_rep2$`treat-4` <- as.numeric(cps_rep2$yearFromSDRFactor == '-4')

cps_rep2$`treat-2` <- as.numeric(cps_rep2$yearFromSDRFactor == '-2')

cps_rep2$`treat0` <- as.numeric(cps_rep2$yearFromSDRFactor == '0')

cps_rep2$`treat2` <- as.numeric(cps_rep2$yearFromSDRFactor == '2')

cps_rep2$`treat4` <- as.numeric(cps_rep2$yearFromSDRFactor == '4')

cps_rep2$`treat6` <- as.numeric(cps_rep2$yearFromSDRFactor == '6')

cps_rep2$`treat8` <- as.numeric(cps_rep2$yearFromSDRFactor == '8')

cps_rep2$`treat10` <- as.numeric(cps_rep2$yearFromSDRFactor == '10')

cps_rep2$`treat12` <- as.numeric(cps_rep2$yearFromSDRFactor == '12')

cps_rep2$`treat14` <- as.numeric(cps_rep2$yearFromSDRFactor == '14')

cps_rep2$`treat16` <- as.numeric(cps_rep2$yearFromSDRFactor == '16')

# Create indicator for inclusion in the regression
cps_rep2$include <- as.numeric(cps_rep2$yearFromSDR <=16 & 
                                 cps_rep2$yearFromSDR >=(-16) & 
                                 cps_rep2$always_treated ==0)
```

# Coefficient plot

```{r}

# By age-group
age_group_levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")

# Store results for each age group
age_group_results <- list()

# For each age group
for (group_level in age_group_levels){
  #Fit a regression to get the per-period tests

  dyn_reg_rep2 <- lm_robust(voted ~ `treat-16` +  `treat-14` +  `treat-12` +  
                              `treat-10`  + `treat-8`  + `treat-6` + `treat-4` +
                              `treat0`  + `treat2`  + `treat4`  + `treat6` +  
                              `treat8` + `treat10` +`treat12`  + `treat14`  + 
                              `treat16` , 
                         fixed_effects = ~ as.factor(year) + as.factor(state),
                         data=cps_exact[
                           cps_exact$age_group==group_level & 
                             (cps_rep2$include == 1| 
                                cps_rep2$yearFromSDRFactor == "Inf") &
                             (cps_exact$yearFromSDR>=(-16) &
                                cps_exact$yearFromSDR<=16) & 
                             !is.na(cps_exact$voted),], 
                         clusters=state, se_type="stata")
  
  #print(dyn_reg_rep2)
  
  # Make into a plotting set
  dyn_plot_2 <- tidy(dyn_reg_rep2) %>% 
    filter(!is.na(estimate)) %>% 
    mutate(term = str_remove_all(term, "`")) %>% 
    mutate(period = as.numeric(str_remove(term, fixed("treat")))) %>% 
    select(period, estimate, conf.low, conf.high)
  
  # Tack on the baseline
  dyn_plot_2 <- rbind(dyn_plot_2, data.frame(period = -2, estimate = 0, 
                                             conf.low = NA, conf.high = NA))
  
  # Label the age_group
  dyn_plot_2$age_group <- group_level
  
  age_group_results[[group_level]] <- dyn_plot_2
  
}

dyn_plot_combined <- bind_rows(age_group_results)

# Coefficient plot
replication_2 <- dyn_plot_combined %>% 
  ggplot(aes(x=as.numeric(period), y = estimate, ymin=conf.low, ymax=conf.high)) + 
  geom_ribbon(fill="grey80") +
  geom_line() +
  facet_wrap(~age_group) +
  geom_vline(xintercept = -0, lty=2) + geom_hline(yintercept = 0, lty=2) +
  xlab("Years Until SDR Implementation") + ylab("Treatment Effect")  + 
  scale_x_continuous(breaks=c(-16,-12,-8,-4,0,4,8,12,16)) +
  scale_y_continuous(breaks=c(-.2,-.15, -0.1,-0.05,0,0.05,0.1,0.15, 0.2), 
                     limits=c(-.2,.2)) +
  theme_classic()

print(replication_2)
ggsave("figs/pretrend_by_age_rep_2.png",replication_2, width = 7, height = 4.3)
ggsave("figs/pretrend_by_age_rep_2.pdf",replication_2, width = 7, height = 4.3)
```

Drop those treated super-late
```{r}

# Coefficient plot
# By age-group
age_group_levels = c("18-24", "25-34", "35-44", "45-54", "55-64", "65+")

# Store results for each age group
age_group_results <- list()

# For each age group
for (group_level in age_group_levels){
  #Fit a regression to get the per-period tests

  dyn_reg_rep3 <- lm_robust(voted ~ `treat-16` +  `treat-14` +  `treat-12` +  
                              `treat-10`  + `treat-8`  + `treat-6` + `treat-4`+ 
                              `treat0`  + `treat2`  + `treat4`  + `treat6` + 
                              `treat8` + `treat10` +`treat12`  + `treat14`  + 
                              `treat16` , 
                         fixed_effects = ~ as.factor(year) + as.factor(state),
                         data=cps_exact[
                           cps_exact$age_group==group_level&
                             (cps_rep2$include == 1|
                                cps_rep2$yearFromSDRFactor == "Inf") &
                             (cps_exact$yearFromSDR>=(-16) &
                                cps_exact$yearFromSDR<=16) & 
                             !is.na(cps_exact$voted),] %>% 
                           filter(!(st %in% c("UT", "HI", "VT"))), 
                         clusters=state, se_type="stata")
  
  #print(dyn_reg_rep3)
  
  # Make into a plotting set
  dyn_plot_3 <- tidy(dyn_reg_rep3) %>% 
    filter(!is.na(estimate)) %>% 
    mutate(term = str_remove_all(term, "`")) %>% 
    mutate(period = as.numeric(str_remove(term, fixed("treat")))) %>% 
    select(period, estimate, conf.low, conf.high)
  
  # Tack on the baseline
  dyn_plot_3 <- rbind(dyn_plot_3, data.frame(
    period = -2, estimate = 0, conf.low = NA, conf.high = NA))
  
  # Label the age_group
  dyn_plot_3$age_group <- group_level
  
  age_group_results[[group_level]] <- dyn_plot_3
  
}

dyn_plot_combined <- bind_rows(age_group_results)

# Coefficient plot
replication_3 <- dyn_plot_combined %>% 
  ggplot(aes(x=as.numeric(period), y = estimate, ymin=conf.low, ymax=conf.high)) + 
  geom_ribbon(fill="grey80") +
  geom_line() +
  facet_wrap(~age_group) +
  geom_vline(xintercept = -0, lty=2) + geom_hline(yintercept = 0, lty=2) +
  xlab("Years Until SDR Implementation") + ylab("Treatment Effect")  + 
  scale_x_continuous(breaks=c(-16,-12,-8,-4,0,4,8,12,16)) +
  scale_y_continuous(breaks=c(-.2,-.15, -0.1,-0.05,0,0.05,0.1,0.15, 0.2), 
                     limits=c(-.2,.2)) +
  theme_classic()

print(replication_3)

ggsave("figs/pretrend_by_age_dropped_late_T.png",replication_3, 
       width = 7, height = 4.3)
ggsave("figs/pretrend_by_age_dropped_late_T.pdf",replication_3, 
       width = 7, height = 4.3)
```

### corrected replication using CS did estimator
- Control group: never treated and yet-to-be-treated
- Base period (placebo): universal (set to -1)

```{r, warning=F, message=F}
######did correct spec#####
for (group_level in age_group_levels){
  
  # Use att_gt from did to retrieve ATTs
  dyn_reg_rep_did2 <- att_gt(yname = "voted",
                        tname = "year",
                        idname = "statefip",
                        gname = "yearSDR_did",
                        xformla = ~1,
                        base_period = "universal",
                        control_group = c("nevertreated", "notyettreated"),
                        allow_unbalanced_panel = T,
       data=cps_exact[cps_exact$age_group==group_level& 
                        (cps_exact$yearFromSDR>=(-16) &
                           cps_exact$yearFromSDR<=16)&
                        !is.na(cps_exact$voted),])

dyn_reg_rep_did2 <- aggte(dyn_reg_rep_did2, type="dynamic", na.rm = T)
  
  #print(dyn_reg_rep_did2)
  
  # Make into a plotting set
  dyn_plot_did2 <- tidy(dyn_reg_rep_did2) %>% 
    filter(!is.na(estimate)) %>% 
    mutate(term = str_remove_all(term, "`")) %>% 
    mutate(period = as.numeric(str_remove_all(term, ("ATT\\(|\\)")))) %>% 
    select(period, estimate, conf.low, conf.high)
  
  # Tack on the baseline
  dyn_plot_did2 <- rbind(dyn_plot_did2, data.frame(period = -2, estimate = 0, 
                                                   conf.low = NA, conf.high = NA))
  
  # Label the age_group
  dyn_plot_did2$age_group <- group_level
  
  age_group_results[[group_level]] <- dyn_plot_did2
  
}

dyn_plot_combined_did2 <- bind_rows(age_group_results)

# Coefficient plot
replication_did2 <- dyn_plot_combined_did2 %>% 
  ggplot(aes(x=as.numeric(period), y = estimate, ymin=conf.low, ymax=conf.high)) + 
  geom_ribbon(fill="grey80") +
  geom_line() +
  facet_wrap(~age_group) +
  geom_vline(xintercept = -0, lty=2) + geom_hline(yintercept = 0, lty=2) +
  xlab("Years Until SDR Implementation") + ylab("Treatment Effect")  + 
  scale_x_continuous(breaks=c(-16,-12,-8,-4,0,4,8,12,16)) +
  scale_y_continuous(breaks=c(-0.2, -.15, -0.1,-0.05,0,0.05,0.1,0.15, 0.2)) +
  theme_classic()

print(replication_did2)

ggsave("figs/pretrend_by_age_did2_universal.png",replication_did2, width = 7, 
       height = 4.3)
ggsave("figs/pretrend_by_age_did2_universal.pdf",replication_did2, width = 7, 
       height = 4.3)

```

### Corrected replication using CS did estimator
- Control group: never treated and yet-to-be-treated
- Base period (placebo): varying (one period before the estimated period)

```{r, warning=F, message=F}

######did correct spec#####
for (group_level in age_group_levels){
  #Fit a regression to get the per-period tests

dyn_reg_rep_did3 <- att_gt(yname = "voted",
                        tname = "year",
                        idname = "statefip",
                        gname = "yearSDR_did",
                        xformla = ~1,
                        base_period = "varying",
                        control_group = c("nevertreated", "notyettreated"),,
                        allow_unbalanced_panel = T,
                        data=cps_exact[
                          cps_exact$age_group==group_level& 
                            (cps_exact$yearFromSDR>=(-16) &
                               cps_exact$yearFromSDR<=16)&
                            !is.na(cps_exact$voted),])

dyn_reg_rep_did3 <- aggte(dyn_reg_rep_did3, type="dynamic", na.rm = T)
  
  #print(dyn_reg_rep_did3)
  
  # Make into a plotting set
  dyn_plot_did3 <- tidy(dyn_reg_rep_did3) %>% filter(!is.na(estimate)) %>% 
    mutate(term = str_remove_all(term, "`")) %>% 
    mutate(period = as.numeric(str_remove_all(term, ("ATT\\(|\\)")))) %>% 
    select(period, estimate, conf.low, conf.high)
  
  # Tack on the baseline
  dyn_plot_did3 <- rbind(dyn_plot_did3, data.frame(
    period = -2, estimate = 0, conf.low = NA, conf.high = NA))
  
  # Label the age_group
  dyn_plot_did3$age_group <- group_level
  
  age_group_results[[group_level]] <- dyn_plot_did3
  
}

dyn_plot_combined_did3 <- bind_rows(age_group_results)

# Coefficient plot
replication_did3 <- dyn_plot_combined_did3 %>% 
  ggplot(aes(x=as.numeric(period), y = estimate, ymin=conf.low, ymax=conf.high)) + 
  geom_ribbon(fill="grey80") +
  geom_line() +
  facet_wrap(~age_group) +
  geom_vline(xintercept = -0, lty=2) + geom_hline(yintercept = 0, lty=2) +
  xlab("Years Until SDR Implementation") + ylab("Treatment Effect")  + 
  scale_x_continuous(breaks=c(-16,-12,-8,-4,0,4,8,12,16)) +
  scale_y_continuous(breaks=c(-0.2, -.15, -0.1,-0.05,0,0.05,0.1,0.15, 0.2)) +
  theme_classic()

print(replication_did3)

ggsave("figs/pretrend_by_age_did3_varying.png",replication_did3, width = 7, 
       height = 4.3)
ggsave("figs/pretrend_by_age_did3_varying.pdf",replication_did3, width = 7, 
       height = 4.3)

```
