---
title: "Week 6: Instrumental Variables"
subtitle: "PLSC 30600 - Causal Inference"
# author: "Anton Strezhnev"
output: 
  xaringan::moon_reader:
    self_contained: true
    css: [default, uchicago_pol_meth.css]
    nature:
      highlightLines: true
      ratio: '16:9'
  
---

# Last two weeks

  - Identification under **conditional ignorability**
    - Treatment assignment is independent of the potential outcomes given observed confounders $\mathbf{X}$
    - "Selection-on-observables"
  - "Selection-on-observables" isn't a testable assumption
    - Relies on theory to decide which $\mathbf{X}$ to include.
    - DAGs can help here.
  - Lots of estimation strategies
    - Stratify with low-dimensional $\mathbf{X}$
    - IPTW to eliminate treatment-covariate relationship, regression to model the outcome-covariate relationship.
    - Matching to reduce model dependence.
      - Or consider more modern flexible modelling techniques for $E[Y_i(d) | X_i]$
      
---

# This week

  - Can we estimate a treatment effect when neither ignorability nor conditional ignorability hold for treatment?
    - Can we get rid of *unobserved* confounding?
  - "Instrumental variables" designs are one way of dealing with this 
--

  - We can identify *some* average of treatment effects if...
    - There *does* exist an ignorable or conditionally ignorable **instrument** which...
    - ...has a monotonic effect on the treatment...
    - ...and has no effect on the outcome *except* through its effect on the treatment.
--
  - What's the average? The "Local Average Treatment Effect"
    - Average effect among those who are *moved* to take treatment by the instrument
    

---

class: title-slide

# Instrumental Variables
$$
  \require{cancel}
$$

---

# Treatment non-compliance

- Often experiments suffer from treatment **non-compliance**
  - Participants randomized to receive a phone call don't pick up.
  - Participants randomized to wear surgical masks choose not to.
--

- New notation!
  - Let $Z_i$ denote whether $i$ is assigned to receive a treatment.
  - Let $D_i$ denote the treatment actually *taken* by an individual. 
--

- Can we just take the simple difference-in-means between $D_i = 1$ and $D_i = 0$?
  - No! Non-compliance affected by other factors which might also affect the outcome.
  - We're stuck with an observational design.
  
---

- Unless...
  
---

# Intent-to-treat effect

- We can first just change the question - instead of the effect of **treatment**, we can make our estimand the effect of being **assigned to treatment**.
- Our estimator for the ITT is just the difference in means between the $Z_i = 1$ and $Z_i = 0$ arms

$$\hat{\tau}_{\text{ITT}} = \hat{E}[Y_i | Z_i = 1] - \hat{E}[Y_i | Z_i = 0]$$

- Identified under randomization of $Z_i$ even if $D_i$ is not randomized.
  - But combines two effects: the actual effect of $D_i$ and the effect of $Z_i$ on $D_i$.

---

# Instrumental variables

```{tikz , echo=F, fig.align='center'}
\usetikzlibrary{shapes}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.symbols}
\usetikzlibrary{shadows}
\usetikzlibrary{fit}
\begin{tikzpicture}
    \node (z1) at (-1,0) {$Z$};
    \node (a1) at (0,0) {$D$};
    \node (l1) at (0.5, 1)
    {$U$};
    \node (y) at
    (1,0) {$Y$};
    \draw[->, >=stealth, thick, dashed] (l1) -- (a1);
    \draw[->, >=stealth, thick, dashed] (l1) --  (y);
    \draw[->, >=stealth, thick] (a1) -- (y);
    \draw[->, >=stealth, thick] (z1) -- (a1);
\end{tikzpicture}
```

- Suppose though that we're interested in the *actual* effect of receiving treatment (the effect of $D_i$). What can we do?



---

# Instrumental variables

- Start by writing down potential outcomes for $D_i$ along with .maroon[joint] potential outcomes of $Y_i$ in terms of $Z_i$ and $D_i$

$$D_i(z) = D_i \text{ if } Z_i = z$$
$$Y_i(d, z) = Y_i \text{ if } D_i = d, Z_i = z$$
--

- Observed treatment $D_i$ is a function of treatment assignment ( $Z_i$ ) - it's a post-treatment quantity (and so has potential outcomes).

---

# Assumptions

1. Randomization of instrument
2. Exclusion restriction
3. Non-zero first-stage relationship
4. Monotonicity

---

# Assumption 1: Randomization

- $Z_i$ is is independent of both sets of potential outcomes (potential outcomes for the treatment and potential outcomes for the outcome).

$$\{D_i(1), D_i(0)\} {\perp \! \! \! \perp} Z_i$$

$$\{Y_i(d, z) \forall d, z\} {\perp \! \! \! \perp} Z_i$$
--

- We can weaken this to conditional ignorability (where $Z_i$ is randomized conditional on $X_i$), which is common in observational settings.
  - But if we don't believe conditional ignorability for the treatment, why would we believe it for the instrument?
- Sufficient to identify the **intent-to-treat (ITT)** effect

---

# Assumption 1: Randomization

```{tikz , echo=F, fig.align='center'}
\usetikzlibrary{shapes}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.symbols}
\usetikzlibrary{shadows}
\usetikzlibrary{fit}
\begin{tikzpicture}
    \node (z1) at (-1,0) {$Z$};
    \node (a1) at (0,0) {$D$};
    \node (l1) at (0.5, 1)
    {$U$};
    \node (y) at
    (1,0) {$Y$};
    \draw[->, >=stealth, thick, dashed] (l1) -- (a1);
    \draw[->, >=stealth, thick, dashed] (l1) --  (y);
    \draw[->, >=stealth, thick] (a1) -- (y);
    \draw[->, >=stealth, thick] (z1) -- (a1);
   \draw[->, >=stealth, red!60, thick, dashed] (l1) -- node[cross out,draw, solid]{} (z1);
\end{tikzpicture}
```

- The randomization assumption eliminates any arrows from $U$ to $Z$.

---

# Assumption 2: Exclusion restriction

- $Z_i$ **only affects** $Y_i$ by way of its effect on $D_i$.
- In other words, if $D_i$ were set at some level $d$, the potential outcome for $Y_i(d, z)$ does not depend on $z$.

$$Y_i(d, z) = Y_i(d, z^{\prime}) \text{ for any } z \neq z^{\prime}$$

--

- **Not a testable assumption!** -- we have to justify this with substantive knowledge.
  - Easiest in the treatment non-compliance case
  - But consider what might happen in a non-blinded situation where respondents knew their treatment assignments.
- "Surprise" factor -- If I told you $Z$ was associated with $Y$, would you think "that's odd"?

---

# Assumption 2: Exclusion restriction

```{tikz , echo=F, fig.align='center'}
\usetikzlibrary{shapes}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.symbols}
\usetikzlibrary{shadows}
\usetikzlibrary{fit}
\begin{tikzpicture}
    \node (z1) at (-1,0) {$Z$};
    \node (a1) at (0,0) {$D$};
    \node (l1) at (0.5, 1)
    {$U$};
    \node (y) at
    (1,0) {$Y$};
    \draw[->, >=stealth, thick, dashed] (l1) -- (a1);
    \draw[->, >=stealth, thick, dashed] (l1) --  (y);
    \draw[->, >=stealth, thick] (a1) -- (y);
    \draw[->, >=stealth, thick] (z1) -- (a1);
    \draw[->, >=stealth, red!60, thick] (z1)  to[bend right = 50] node[cross out, -, draw, red!60, thick, midway] {\vspace{3em}} node[red!60, midway, below] {}(y);
\end{tikzpicture}
```


- The exclusion restriction eliminates any causal paths from $Z$ to $Y$ **except** for $Z \to D \to Y$.

---

# Assumption 3: Non-zero first stage

- $Z_i$ has an effect on $D_i$

$$E[D_i(1) - D_i(0)] \neq 0$$
--

- Seems trivial, but we need this to make the estimator work.
- Magnitude matters for estimator performance - a "weak" first-stage $\leadsto$ heavily biased IV estimator
  - IV estimators are *consistent* but not *unbiased*.

---

# Assumption 3: Non-zero first stage

```{tikz , echo=F, fig.align='center'}
\usetikzlibrary{shapes}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.symbols}
\usetikzlibrary{shadows}
\usetikzlibrary{fit}
\begin{tikzpicture}
    \node (z1) at (-1,0) {$Z$};
    \node (a1) at (0,0) {$D$};
    \node (l1) at (0.5, 1)
    {$U$};
    \node (y) at
    (1,0) {$Y$};
    \draw[->, >=stealth, thick, dashed] (l1) -- (a1);
    \draw[->, >=stealth, thick, dashed] (l1) --  (y);
    \draw[->, >=stealth, thick] (a1) -- (y);
    \draw[->, >=stealth, thick, red!60, thick] (z1) -- (a1);
\end{tikzpicture}
```

- The non-zero first stage assumption requires a path from $Z$ to $D$.

---

# Assumption 4: Monotonicity

- $Z_i$'s effect on $D_i$ only goes in one direction **at the individual level**

$$D_i(1) - D_i(0) \ge 0$$

- If it goes the other way, we can always flip the direction of the treatment to make this hold 
  - The key is that the instrument does not have a positive effect on $D_i$ for some units and a negative effect for others.
- **Not a testable assumption**

---

# Assumption 4: Monotonicity

- In binary instrument/binary treatment world, this is sometimes called a "no defiers" assumption.

Stratum              | $D_i(1)$          |  $D_i(0)$ | 
:-------------------:|:-----------------:|:---------:|
"Always-takers"      |  $1$              |  $1$      |  
"Never-takers"       |  $0$              |  $0$      |  
"Compliers"          |  $1$              |  $0$      | 
"Defiers"            |  $0$              |  $1$      |

--

- Under no defiers, every unit with $D_i = 1$ and $Z_i = 0$ is an always-taker, every unit with $D_i = 0$ and $Z_i =1$ is a never-taker.

---

# Assumption 4: Monotonicity

```{tikz , echo=F, fig.align='center'}
\usetikzlibrary{shapes}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{shapes.misc}
\usetikzlibrary{shapes.symbols}
\usetikzlibrary{shadows}
\usetikzlibrary{fit}
\begin{tikzpicture}
    \node (z1) at (-1,0) {$Z$};
    \node (a1) at (0,0) {$D$};
    \node (l1) at (0.5, 1)
    {$U$};
    \node (y) at
    (1,0) {$Y$};
    \draw[->, >=stealth, thick, dashed] (l1) -- (a1);
    \draw[->, >=stealth, thick, dashed] (l1) --  (y);
    \draw[->, >=stealth, thick] (a1) -- (y);
    \draw[->, >=stealth, thick, red!60, thick] (z1) -- (a1);
\end{tikzpicture}
```

- Can't represent the monotonicity assumption in a DAG - it's an assumption about the form of the relationship between $Z$ and $D$.

---

# Interpreting the IV estimand

- The classic IV estimand with one instrument is a ratio of sample covariances.

$$\tau_{\text{IV}} = \frac{Cov(Y, Z)}{Cov(D, Z)}$$

- With a binary instrument, this is sometimes called the "Wald" estimand - a ratio of differences in means

$$\tau_{\text{IV}} = \frac{E[Y_i | Z_i = 1] - E[Y_i | Z_i = 0]}{E[D_i | Z_i = 1] - E[D_i | Z_i = 0]}$$

---

# Interpreting the IV estimand

- What does the Wald estimand correspond to in terms of causal effects?

$$\tau_{\text{IV}} = \frac{E[Y_i | Z_i = 1] - E[Y_i | Z_i = 0]}{E[D_i | Z_i = 1] - E[D_i | Z_i = 0]}$$

--

- Under our identification assumptions:
  - The numerator is the ITT
  - The denominator is the first-stage effect
  
---

# Interpreting the IV estimand

- Let's decompose the denominator first - under randomization:
$$\begin{align*}
E[D_i | Z_i = 1] - E[D_i | Z_i = 0] &= E[D_i(1) | Z_i = 1] - E[D_i(0) | Z_i = 0]\\
&= E[D_i(1)] - E[D_i(0)]\\
&= E[D_i(1) - D_i(0)]
\end{align*}$$

--

- With binary treatment/binary instrument, we can use law of total expectation to decompose by principal stratum

$$E[D_i(1) - D_i(0)] = E[D_i(1) - D_i(0) | D_i(1) = D_i(0)] \times P(D_i(1) = D_i(0)) + \\
E[D_i(1) - D_i(0) | D_i(1) > D_i(0)] \times P(D_i(1) > D_i(0)) + \\
E[D_i(1) - D_i(0) | D_i(1) < D_i(0)] \times P(D_i(1) < D_i(0))$$

--

- The first term is $0$
- And by no defiers, the last term is $0$ since $P(D_i(1) < D_i(0)) = 0$

$$E[D_i(1) - D_i(0)] = Pr(D_i(1) > D_i(0))$$


---

# Interpreting the IV estimand

- Next, the numerator (the ITT). Under the exclusion restriction and randomization:

$$E[Y_i | Z_i = 1] = E\bigg[Y_i(0) + \bigg(Y_i(1) - Y_i(0)\bigg)D_i(1)\bigg]$$
$$E[Y_i | Z_i = 0] = E\bigg[Y_i(0) + \bigg(Y_i(1) - Y_i(0)\bigg)D_i(0)\bigg]$$
--

- The difference (with some algebra) is

$$E[Y_i | Z_i = 1] - E[Y_i | Z_i = 0]  = E\bigg[\bigg(Y_i(1) - Y_i(0)\bigg) \times \bigg(D_i(1) - D_i(0)\bigg)\bigg]$$

---

# Interpreting the IV estimand

- Conditioning on the principal strata:

$$= E\bigg[(Y_i(1) - Y_i(0)) \times (0) | (D_i(1) = D_i(0))\bigg] \times P(D_i(1) = D_i(0)) + \\
 E\bigg[(Y_i(1) - Y_i(0)) \times (1) | (D_i(1) > D_i(0))\bigg] \times  P(D_i(1) > D_i(0)) + \\
 E\bigg[(Y_i(1) - Y_i(0)) \times (-1) | (D_i(1) < D_i(0))\bigg] \times P(D_i(1) < D_i(0))$$
 
 - Again, first term is zero because $D_i(1) - D_i(0) = 0$, third is zero by "no defiers" and we have
 
$$E[Y_i | Z_i = 1] - E[Y_i | Z_i = 0] =  E\bigg[Y_i(1) - Y_i(0) | D_i(1) > D_i(0)\bigg] \times  P(D_i(1) > D_i(0))$$

- The ITT is the product of a conditional average treatment effect and the proportion of compliers.

---

# The LATE Theorem

- The IV estimand, under our identification assumptions, is a Local Average Treatment Effect (LATE):

$$\frac{E[Y_i | Z_i = 1] - E[Y_i | Z_i = 0]}{E[D_i | Z_i = 1] - E[D_i | Z_i = 0]} = E[Y_i(1) - Y_i(0) | D_i(1) > D_i(0)]$$

- The LATE is a conditional average treatment effect within the *subpopulation* of **compliers**
- If treatment effects are constant, we can generalize this to the whole sample.
  - But if effects are heterogeneous, we are not necessarily getting a "representative" treatment effect.
  
---

# Better LATE than never?

- How should we interpret the LATE?  
  - It's not necessarily the quantity we care about - we care about the effect of the treatment in the entire sample.
--

- Compliers are those compelled to take treatment by our encouragement. Would estimates generalize to those who are less encourageable?
  - The LATE is design-specific. If we came up with a different instrument, that changes the population on which we're estimating an effect!
  - What can we do? 
    - We could describe the distribution of covariates among compliers vs. the population as a whole (Abadie's kappa-weighting).

---


# Example: The effect of media on voting

- Gerber, Karlan and Bergan (2009, AEJ:AE) estimate the effect of reading the Washington Post (or Washington Times) on political attitudes and voting behavior.
  - $Z_i$: Random assignment to receive a free subscription to the Washington Post
  - $D_i$: Actually subscribing to the Washington Post (as measured by a post-encouragement survey)
  - $Y_i$: 2005 Turnout (measured in the survey)

--
- Assumptions:
  - Assignment to get the free subscription offer is ignorable/exogenous
  - Getting the free subscription offer affects actual subscriptions (non-zero first stage)
  - No one would subscribe to the Post if they *didn't* receive the offer but not subscribe if they *did*. (monotonicity/no defiers)
  - Assignment to get the free subscription offer doesn't affect voting *except through* actually subscribing to the Post (exclusion restriction) 

---

# Example: The effect of media on voting

- First, subset the data to WaPo or control observations that completed the follow-up survey

```{r, echo=F, message=F, warning=F}
library(tidyverse)
library(haven)
library(estimatr)
library(ivmodel)
options(digits=3)
```

```{r}
green <- read_dta("assets/publicdata.dta")
wapost <- green %>% filter(treatment != "TIMES"&!is.na(getpost)&!is.na(voted))
```


---

# Example: The effect of media on voting

- Is there a first-stage effect?

```{r}
lm_robust(getpost ~ post, data=wapost)
```

- About 34 percent of the sample is a "complier" - quite substantial!
 
---

# Example: The effect of media on voting

- Is there an ITT?

```{r}
lm_robust(voted ~ post, data=wapost)
```

- ITT is essentially zero.
 
---

# Example: The effect of media on voting

- Compare with the naive OLS estimate

```{r}
lm_robust(voted ~ getpost, data=wapost)
```

- Post subscribers are 6pp more likely to vote in the 2005 VA gubernatorial election. 
  - But is this causal No!
 
---


# Example: The effect of media on voting

- Let's estimate the LATE using the Wald estimator 

```{r}
(mean(wapost$voted[wapost$post == 1]) - mean(wapost$voted[wapost$post == 0]))/(mean(wapost$getpost[wapost$post == 1]) - mean(wapost$getpost[wapost$post == 0]))
```

- Equivalent to a ratio of regression coefficients
```{r}
coef(lm_robust(voted ~ post, data=wapost))[2]/coef(lm_robust(getpost ~ post, data=wapost))[2]
```

 
---

# Example: The effect of media on voting

- We'll talk about inference later, but take note: the SE for the LATE can be much larger than the SE for the ITT

```{r}
iv_robust(voted ~ getpost | post, data=wapost)
```

---

# IV in observational studies

- Most applications of IV are not treatment non-compliance.
- But all follow the same underlying logic.
  - Treatment of interest is not randomized...but there exists a real or "natural" experiment that is.
  - And this natural experiment affects the outcome only through its affect on the treatment of interest.
--

- Examples:
  - Angrist (1990) - Vietnam draft lottery number as an instrument for the effect of military service on income.
  - Angrist and Krueger (1991) - Birth quarter as an instrument for education's effect on income.
  - Acemoglu et. al. (2001) - European settler mortality as an instrument for the effect of institutional quality on GDP per capita.
  - Kern & Hainmueller (2009) - West German TV signal strength as an instrument for the effect of watching West German TV on support for the East German regime

--
- Challenges
  - Exogeneity/ignorability isn't guaranteed
  - If an instrument has a large effect on your treatment of interest, it probably has an effect on other stuff that could affect the outcome as well (violating the exclusion restriction)
  
---

# Discussion: The rainfall instrument

- **Miguel, Satyanth, and Sergenti (2004, JPE)** look at the effect of .maroon[economic growth] on .green[civil conflict] in 41 African countries.
  - Growth and conflict are confounded (e.g. by political institutions).
  - Instrument for GDP growth using the annual change in rainfall -- for heavily agrarian countries, rainfall fluctuations determine crop yields which are a large component of GDP.
  - Observe that changes in rainfall are associated with changes in GDP and negative GDP shocks (instrumented by rainfall) increase civil conflict.
--

- Does this satisfy the IV identification assumptions?
  - Exogeneity? Is rainfall as-good-as randomly assigned?
  - Monotonicity? Do positive rainfall shocks *strictly* boost GDP per capita?
  - Exclusion restriction? Is rainfall's effect transmitted only through the mechanism the authors define?

---

class: title-slide

# Estimation and inference for IV


---

# IV with constant effects

- Let's consider a linear model for the potential outcomes

$$Y_i(d) = \alpha + \tau d + \gamma U_i + \eta_i$$

--

- If we could control for $U_i$, we could estimate the regression to get an estimate of $\tau$

$$Y_i = \alpha + \tau D_i + \gamma U_i + \eta_i$$

- But we can't - and regressing $Y_i$ on $D_i$ alone will not give a consistent estimator of $\tau$ since $Cov(\gamma U_i + \eta_i, D_i) \neq 0$

---

# IV with constant effects

- Suppose $Z_i$ is an instrument that is exogeneous and satisfies the exclusion restriction

$$Cov(\gamma U_i + \eta_i, Z_i) = 0$$
--

- Then, we can identify $\tau$

$$\begin{align*}
Cov(Y_i, Z_i) &= Cov(\alpha + \tau D_i + \gamma U_i + \eta_i, Z_i)\\
&= Cov(\alpha, Z_i) + Cov(\tau D_i, Z_i) + Cov(\gamma U_i + \eta_i, Z_i)\\
&= \tau Cov( D_i, Z_i)
\end{align*}$$

--

- Which gives us our IV estimand

$$\tau = \frac{Cov(Y_i, Z_i)}{Cov(D_i, Z_i)}$$

---

# IV estimator

- We can estimate $\tau$ by plugging in the sample quantities.

$$\hat{\tau_{\text{IV}}} = \frac{\widehat{Cov(Y_i, Z_i)}}{\widehat{Cov(D_i, Z_i)}}$$
--

- This also can be written as a ratio of two regression coefficients

$$\hat{\tau_{\text{IV}}} = \frac{\widehat{Cov(Y_i, Z_i)}/\widehat{Var(Z_i)}}{\widehat{Cov(D_i, Z_i)}/\widehat{Var(Z_i)}}$$

- Denominator: "First stage": Regression of $D_i$ on $Z_i$
- Numerator: "Reduced form": Regression of $Y_i$ on $Z_i$

---

# 2SLS - Including covariates

- What if ignorability of $Z_i$ only holds conditional on $X_i$ -- or we want to include $X_i$ as predictors to improve precision.
- We'll assume a particular structure for the outcome and treatment models

$$Y_i = X_i^{\prime}\beta + \tau D_i + \epsilon_i$$
$$D_i = X_i^{\prime}\alpha + \gamma Z_i + \nu_i$$

- Assume the $X_i$ are exogenous but not excluded (appear in both equations). $D_i$ is still endogenous so we can't get the treatment effect by just regressing outcome on treatment and covariates.

- Can we get an expression for $Y_i$ in the form of $Z_i$ alone?

---

# 2SLS - Including covariates

- Substitute in for $D_i$

$$\begin{align*}
Y_i &= X_i^{\prime}\beta + \tau\left[X_i^{\prime}\alpha + \gamma Z_i + \nu_i\right] + \epsilon_i\\
 &= X_i^{\prime}\beta + \tau\left[X_i^{\prime}\alpha + \gamma Z_i\right] + [\tau\nu_i + \epsilon_i]\\
  &= X_i^{\prime}\beta + \tau E[D_i | X_i, Z_i] + \epsilon_i^{*}\\
\end{align*}$$

--

- We can identify $\tau$ by regressing $Y_i$ on $X_i$ and the **fitted values** from a regression of $D_i$ on $X_i$ and the instrument $Z_i$.
- **Intuition** -- we want to only use the variation in $D_i$ that is driven by the **exogenous** factor $Z_i$. 

---

# 2SLS - Including covariates

- You can also still get the ratio form of the IV estimator

$$\begin{align*}
Y_i &= X_i^{\prime}\beta + \tau\left[X_i^{\prime}\alpha + \gamma Z_i + \nu_i\right] + \epsilon_i\\
 &= X_i^{\prime}(\beta + \tau\alpha) + \tau \gamma Z_i + [\tau\nu_i + \epsilon_i]\\
\end{align*}$$

- The coefficient on $Z_i$ in the reduced form regression is $\tau \gamma$ 
- The coefficient on $Z_i$ in the first-stage is $\gamma$.
- So the ratio of the reduced form to the first-stage regression is $\tau$.

---

# Two-stage least squares

- **First stage** - Regress $D_i$ on $X_i$ and $Z_i$. Get the fitted values $\hat{D_i}$

$$\hat{D_i} =  X_i^{\prime}\hat{\alpha} + \hat{\gamma} Z_i  $$

- **Second stage** - Regress $Y_i$ on $X_i$ and fitted values $\hat{D_i}$

$$\hat{Y_i} = X_i^{\prime}\hat{\beta} + \hat{\tau} \hat{D_i}$$

--

- The coefficient on the fitted values is the IV estimate 
  - But, the standard errors will be wrong - Why?


---

# General 2SLS

- A more general matrix-algebra-y way of representing 2SLS is with the linear model

$$\mathbf{Y} = \mathbf{X}\beta + \epsilon$$
where $\mathbf{X}$ is an $n$ by $K$ matrix of endogenous variables and exogenous covariates.

--

- $\mathbf{Z}$ is an $n$ by $L$ matrix of exogenous instruments and exogenous covariates.

--

- Key assumptions:
  - $E[\epsilon_i | \mathbf{Z}] = 0$ (instruments are exogenous)
  - $\mathbf{Z}$ is excluded from the outcome model

---

# General 2SLS

- We define a projection matrix $\mathbf{P}_{\mathbf{z}} = \mathbf{Z}(\mathbf{Z}^{\prime}\mathbf{Z})^{-1}\mathbf{Z}^{\prime}$
- The "fitted values" from the first stage can be written as 

$$\hat{\mathbf{X}} = \mathbf{P}_{\mathbf{z}} \mathbf{X}$$
- Two features of the projection matrix: $\mathbf{P}_\mathbf{z}^{\prime} = \mathbf{P}_\mathbf{z}$ and $\mathbf{P}_\mathbf{z}\mathbf{P}_\mathbf{z} = \mathbf{P}_\mathbf{z}$
--

- So our 2SLS IV estimator can be written as 

$$\begin{align*}
\hat{\beta}_{\text{IV}} &= (\hat{\mathbf{X}}^{\prime}\hat{\mathbf{X}})^{-1}\hat{\mathbf{X}}^{\prime}\mathbf{Y}\\
&=  ((\mathbf{P}_{\mathbf{z}} \mathbf{X})^{\prime}\mathbf{P}_{\mathbf{z}} \mathbf{X})^{-1}(\mathbf{P}_{\mathbf{z}} \mathbf{X})^{\prime}\mathbf{Y}\\
&=  (\mathbf{X}^{\prime}\mathbf{P}_{\mathbf{z}} \mathbf{X})^{-1}\mathbf{X}^{\prime}\mathbf{P}_{\mathbf{z}}\mathbf{Y}
\end{align*}$$

- We have an expression for our IV estimator $\hat{\beta}_{\text{IV}}$ purely in terms of $X$ and $Z$

---

# Variance estimation

- Note that the residuals are $\hat{\epsilon} = \mathbf{Y} - \mathbf{X}\hat{\beta}_{\text{IV}}$. We can plug these in to our usual sandwich estimator for heteroskedasticity-robust SEs

- Our sandwich estimator can be written as

$$\widehat{Var(\hat{\beta}_{\text{IV}})} = (\mathbf{X}^{\prime}\mathbf{P}_{\mathbf{z}} \mathbf{X})^{-1}\mathbf{X}^{\prime}\mathbf{P}_{\mathbf{z}}[\text{diag}(\hat{\epsilon}_i^2)]\mathbf{P}_{\mathbf{z}} \mathbf{X}(\mathbf{X}^{\prime}\mathbf{P}_{\mathbf{z}} \mathbf{X})^{-1}$$

- This is how any conventional 2SLS routine in R will calculate the standard errors.

---

# Illustrating 2SLS

- Recall our Gerber, Karlan and Bergan (2009, AEJ:AE) experiment
  - $Z_i$: Random assignment to receive a free subscription to the Washington Post
  - $D_i$: Actually subscribing to the Washington Post (as measured by a post-encouragement survey)
  - $Y_i$: 2005 Turnout (measured in the survey)
  - $X_i$: Gender, Age
  
- Let's load and subset

```{r}
green <- read_dta("assets/publicdata.dta")
wapost <- green %>% filter(treatment != "TIMES"&!is.na(getpost)&!is.na(voted)&!is.na(Bfemale)&!is.na(reportedage))
```

---

# Illustrating 2SLS

- Our first stage regresses subscription on assignment + covariates

```{r}
first_stage <- lm_robust(getpost ~ post + Bfemale + reportedage , data= wapost)
summary(first_stage)
```

---

# Illustrating 2SLS

- Let's actually run 2SLS - I like two routines: `iv_robust` in `estimatr` (does 2SLS with robust SEs) and `ivmodel` in `ivmodel` (does robust 2SLS *and* weak-instrument robust tests + other diagnostics)

```{r}
wapo_2sls <- iv_robust(voted ~ getpost  + Bfemale + reportedage | post + Bfemale + reportedage, data= wapost)
summary(wapo_2sls)
```
---

# Illustrating 2SLS

```{r}
wapo_2sls2 <- ivmodelFormula(voted ~ getpost  + Bfemale + reportedage | post + Bfemale + reportedage, data= wapost, heteroSE=T)
summary(wapo_2sls2)
```

---

# The weak instrument problem

- Our ratio estimator is consistent 

$$\hat{\tau}_{\text{IV}} = \frac{\widehat{Cov(Y_i, Z_i)}}{\widehat{Cov(D_i, Z_i)}} \overset{p}{\to} \tau + \frac{Cov(U_i, Z_i)}{Cov(D_i, Z_i)}$$

- Under exogeneity $Cov(Z_i, U_i)$ is zero.

--

- However, when there are small violations of exogeneity, a weak instrument will amplify them.
- More generally, with a weak instrument, our t-ratio hypothesis tests assuming asymptotic normality will have **incorrect** type-1 error rates.
  - Why? Distributions of ratios are poorly behaved.
  
---

# The weak instrument problem

- Let's use a simulation to see how bad the bias can be in IV versus just a simple OLS regression of outcome on treatment under unobserved confounding.
--

- Let $U_i \sim \mathcal{N}(0, 1)$ be an unobserved confounder. $Z_i \sim \text{Bern}(.5)$ is an **exogenous** instrument.
--

- The probability of treatment is modeled via a logit
$$\text{log}\bigg(\frac{P(D_i = 1 | Z_i, U_i)}{1-P(D_i = 1 | Z_i, U_i)}\bigg) =  \gamma Z_i + U_i$$

- $\gamma$ here captures the relationship between the exogenous instrument $Z_i$ and the treatment
--

- The outcome is a function of $U$ and a mean zero error term $\epsilon_i$ only, so the true treatment effect is $0$

$$Y_i = U_i + \epsilon_i$$

---

# The weak instrument problem

- Let's see how the Wald estimator performs when we have a pretty large effect of $Z_i$ on $D_i$: $\gamma = 2$ and $N = 1000$

```{r, echo=F}
library(ivmodel)
set.seed(60639)

nIter <- 1000
naive <- rep(NA,nIter)
IV <- rep(NA, nIter)
firststage <- rep(NA, nIter)
firststageF <- rep(NA,nIter)
  
for (i in 1:nIter){
  N <- 1000
  U <- rnorm(N)
  Z <- rbinom(N, 1, .5)
  FS <- 2
  prD <- 1/(1 + exp(-(FS*Z + U)))
  D <- rbinom(N, 1, prD)
  Y <- U + rnorm(N)
  naive[i] <- coef(lm(Y ~ D))[2]
  IV[i] <- coef(lm(Y ~ Z))[2]/coef(lm(D ~ Z))[2]
  firststage[i] <- coef(lm(D~Z))[2]
  firststageF[i] <- summary(lm(D~Z))$fstatistic[1]
}
```

```{r}

## First stage effect
mean(firststage)

## F-statistic from the first stage
mean(firststageF)

## Bias of the naive OLS Y ~ X
mean(naive)

## Bias of IV
mean(IV)

```

---

# The weak instrument problem

- Sampling distribution of the naive OLS estimator

```{r, echo=F, fig.align="center"}
hist(naive, main="OLS")
abline(v=mean(naive), col="blue", lty=2)
abline(v=0, col="red", lty=2)
```

---

# The weak instrument problem

- Sampling distribution of the IV estimator

```{r, echo=F, fig.align="center"}
hist(IV, main="Instrumental Variables")
abline(v=mean(IV), col="blue", lty=2)
abline(v=0, col="red", lty=2)
```

---


# The weak instrument problem

- Now, what happens when our instrument is weak: $\gamma = .2$ and $N = 1000$

```{r, echo=F}
library(ivmodel)
set.seed(60639)

nIter <- 1000
naive <- rep(NA,nIter)
IV <- rep(NA, nIter)
firststage <- rep(NA, nIter)
firststageF <- rep(NA,nIter)
  
for (i in 1:nIter){
  N <- 1000
  U <- rnorm(N)
  Z <- rbinom(N, 1, .5)
  FS <- .2
  prD <- 1/(1 + exp(-(FS*Z + U)))
  D <- rbinom(N, 1, prD)
  Y <- U + rnorm(N)
  naive[i] <- coef(lm(Y ~ D))[2]
  IV[i] <- coef(lm(Y ~ Z))[2]/coef(lm(D ~ Z))[2]
  firststage[i] <- coef(lm(D~Z))[2]
  firststageF[i] <- summary(lm(D~Z))$fstatistic[1]
}
```

```{r}

## First stage effect
mean(firststage)

## F-statistic from the first stage
mean(firststageF)

## Bias of the naive OLS Y ~ X
mean(naive)

## Bias of IV
mean(IV)

```

---

# The weak instrument problem

- Sampling distribution of the naive OLS estimator

```{r, echo=F, fig.align="center"}
hist(naive, main="OLS")
abline(v=mean(naive), col="blue", lty=2)
abline(v=0, col="red", lty=2)
```

---

# The weak instrument problem

- Sampling distribution of the IV estimator

```{r, echo=F, fig.align="center"}
hist(IV, main="Instrumental Variables")
abline(v=mean(IV), col="blue", lty=2)
abline(v=0, col="red", lty=2)
```

---

# The weak instrument problem

- When is an instrument too weak?
--

- **Classic result:** Staiger and Stock (1997), Stock and Yogo (2005) use first stage F-statistic thresholds
  - $\leadsto$ heuristic of first-stage F-statistic below 10.
- **Recently:** Lee, Moreira, McCrary, Porter (2020) -- If we want to use the F-statistic as a screen then we actually need $F > 104.7$
--

- Suggestions:
  - **Anderson-Rubin (1949)** test statistic in addition to 2SLS t-ratio inference (with a single instrument) - correct coverage/rejection levels even under weak instruments.
  - Angrist and Pischke (2009) - with "just-identified" IV (number of instruments = number of endogenous variables) bias is usually overwhelmed by the large standard errors.

---

# Conclusion

- Instrumental variables lets us leverage *alternative* sources of randomness to learn about an otherwise confounded causal relationship.
--

- An instrument:
  - Affects treatment
  - Doesn't affect the outcome except through treatment
  - Is ignorable w.r.t the outcome.
--

- LATE theorem: The IV estimand is the ATE among those who would take treatment due to the instrument.
  - With continuous treatment/instrument - a weighted average of LATEs (Angrist and Imbens, 1995)
  - With covariates - a weighted average of covariate-specific LATEs
  - But be careful with this interpretation when the model is not fully saturated (Słoczyński, 2022)

--

- Statistical inference is tricky
  - Beware weak instruments - typical large-sample asymptotics do poorly when instruments are irrelevant.
  - Consider weak-instrument robust tests (Anderson-Rubin)

---